WI FastPlants: Heritability of number of cotelydons

### 2020-07-18 cotyledon-data-analysis.Rmd
updated Rmd with the new data; added some notes about the shiny app

@@

# Data visualization

@@

# Statistical tests

@@

>>>>

# Initial data: `Polycot-data-from-NY.csv`

@@

## Data visualization

@@

## Statistical tests

@@

# New data: `Polycot-data-2020.csv`

Reading in the data (had to remove manually the '#' sign from the column and save as csv file):
```{r}
dat = read.table("data/Polycot-data-2020.csv", sep=",",header=TRUE)
str(dat)
head(dat)
summary(dat)
dat = within(dat, Generation<-factor(Generation, levels=c("P","O")))
```

### Data visualization

```{r, echo=FALSE}
p1 = ggplot(dat, aes(Generation, fill=Generation))+geom_bar()+
  theme(
        plot.title = element_text(hjust=0.5, size=rel(1.8)),
        axis.title.x = element_text(size=rel(1.8)),
        axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
        axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")##,
        )
p1
```

```{r, echo=FALSE}
p3 = ggplot(dat, aes(y = Generation, x = Cotyledons, fill = Generation)) +
  geom_jitter(pch = 21, alpha=0.3, width=0.2)+
  theme(
        plot.title = element_text(hjust=0.5, size=rel(1.8)),
        axis.title.x = element_text(size=rel(1.8)),
        axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
        axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")##,
        )
p3
```

```{r, echo=FALSE}
p3.2 = ggplot(dat, aes(x = Generation, y = Cotyledons, fill = Generation)) +
  geom_jitter(pch = 21, alpha=0.3, height=0.2)+
    geom_boxplot(outlier.size = 0, alpha=0.1) +
  theme(
        plot.title = element_text(hjust=0.5, size=rel(1.8)),
        axis.title.x = element_text(size=rel(1.8)),
        axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
        axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")##,
        )
p3.2
```

```{r, echo=FALSE}
p4 = ggplot(dat, aes(x=Generation, y=Cotyledons, fill=Generation))+geom_violin(alpha=0.5)+
  ylim(c(1,6))+
  theme(
        plot.title = element_text(hjust=0.5, size=rel(1.8)),
        axis.title.x = element_text(size=rel(1.8)),
        axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
        axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")##,
        )

p4
```

```{r, echo=FALSE}
p4.2 = ggplot(dat, aes(x=Generation, y=Cotyledons, fill=Generation))+geom_violin(alpha=1)+
    geom_jitter(pch = 21, alpha=0.1)+
  ylim(c(1,6))+
  theme(
        plot.title = element_text(hjust=0.5, size=rel(1.8)),
        axis.title.x = element_text(size=rel(1.8)),
        axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
        axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")##,
        )

## Statistical tests

- Population 1: Parents
- Population 2: Offspring

```{r, echo=FALSE}
q1 = ggplot(dat, aes(Cotyledons, fill=Generation))+geom_density(alpha=0.25)+
  theme(
        plot.title = element_text(hjust=0.5, size=rel(1.8)),
        axis.title.x = element_text(size=rel(1.8)),
        axis.title.y = element_text(size=rel(1.8), angle=90, vjust=0.5, hjust=0.5),
        axis.text.x = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        axis.text.y = element_text(colour="grey", size=rel(1.5), angle=0, hjust=.5, vjust=.5, face="plain"),
        panel.background = element_blank(),
        axis.line = element_line(colour = "grey")##,
        )
q1
```

```{r}
x = dat$Cotyledons[dat$Generation == "P"]
y = dat$Cotyledons[dat$Generation == "O"]
```

But let's do a Levene test:
```{r}
leveneTest(Cotyledons ~ Generation, dat, center=mean)
```

The p-value is less than 0.05, so we reject the null hypothesis of equal variances.

The Levene test also assumes Normality. So, we can run the alternative Fligner-Killeen test with the same conclusion (equal variances):
```{r}
fligner.test(Cotyledons ~ Generation, dat)
```

- Normality: not met
- Equality of variances: not met

If we ignore the violations and we run a standard t-test, we get:
```{r}
t.test(x,y, var.equal=TRUE)
```

The p-value is less than 0.05, so we reject the null hypothesis of equality of means. This implies that we have evidence that the number of cotyledons is significantly larger in the offspring than in the parents.

We can also run the Welch version of t test that does not assume equal variances (this is the default in R):

```{r}
t.test(x,y)
```

```{r}
wilcox.test(Cotyledons ~ Generation, dat)
```

The p-value less than 0.05 allows us to also reject the null hypothesis of equality of distributions which provides evidence that the number of cotyledons on the offspring group is significantly higher than in the parent group.
